\LARGE \bf Mean Estimation from One-Bit Measurements

### 2019-09-30 appendices.tex
Trying to understand proofs

>>>>

\subsection{Fast convergence of uniform estimators under bit constraints}
\label{sec:uniform-weirdos}

Here we consider the uniform distribution as our location family,
demonstrating that in the adaptive setting~\eqref{item:adaptive} or even the
one-step adaptive setting~(\ref{item:one-step-adaptive}'), constrained
estimators can attain rates faster than the $1 / \sqrt{n}$ rates regular
estimands allow.  Indeed, define $c(x) = -\log 2$ for $x \in [-1, 1]$ and
$c(x) = -\infty$ for $x \not \in [-1, 1]$. Then $f(x) = e^{-c(x)}$ is
log-concave and symmetric, and we may consider the location family with
densities $f(x - \theta)$. For notational simplicity, we assume we have
a sample of size $2n$. We provide a proof sketch that
there is a one-step adaptive estimator $\theta_n$ such that
\begin{equation}
  \label{eqn:uniforms-are-easy}
  \sup_{|\theta| \le \log n}
  P_\theta\left(|\theta_n - \theta| \ge \frac{8 \log n}{n^{3/4}}\right)
  \le \frac{2}{n^2}.
\end{equation}
for all large $n$,
and so (by the Borel-Cantelli lemmas), for any $\theta \in \R$ we have
$P_\theta(|\theta_n - \theta| \le \sqrt{2 \log n} / n^{3/4} ~
\mbox{eventually}) = 1$. This is of course faster than the $1/\sqrt{n}$
rates we prove throughout.

To prove inequality~\eqref{eqn:uniforms-are-easy}, we proceed in two steps,
both quite similar.
First, we define an initial estimator $\theta\init_n$.
Let $\epsilon > 0$, which we will determine presently, though we will
take $n \epsilon \to \infty$ as $n \to \infty$, so that we may assume
w.l.o.g.\ that $\theta \in [-n\epsilon/2, n \epsilon/2]$. Take the interval
$[-n\epsilon, n\epsilon]$, and construct
$m$ thresholds at intervals of size $2 n \epsilon / m$; let
the $j$th such threshold be
\begin{equation*}
  t_j \defeq -n \epsilon + \frac{2 n (j-1) \epsilon}{m}
\end{equation*}
Then we ``assign'' observations to each pair of thresholds, so that
threshold $j$ corresponds to observations $I_j \defeq \{\frac{n (j - 1)}{m}
+ 1, \ldots, \frac{n j}{m}\}$, of which there are $n/m$.  For each index $i
\in I_j$, we set
\begin{equation*}
  B_i = \begin{cases}
    1 & \mbox{if~} X_i \ge t_j \\
    0 & \mbox{otherwise}.
  \end{cases}
\end{equation*}
Then we simply set $\theta_n\init$ to be the maximal threshold for which
$B_i = 0$ for all observations $X_i$ corresponding to that threshold.

Let us now consider the probability that $\theta_n\init$ is substantially
wrong. For notational simplicity, let $U_i = (1 + X_i - \theta) / 2$, so that
the $U_i$ are uniform on $[0, 1]$.
First, note that we always have $\theta_n\init \ge \theta
- \frac{2 n \epsilon}{m}$, because no observations will be below
the appropriate threshold. Let $j\opt$ be the smallest index $j$
for which $\theta \le t_{j\opt}$, and consider
the index sets $I_{j\opt}$, $I_{j\opt + 1}$, and so on.
The event $\theta\init_n \ge t_{j\opt} + \frac{2 n \epsilon}{m}$ may
occur only if for each of the $n/m$ observations in the set
$I_{j\opt + 1}$, we have $U_i \ge \frac{n \epsilon}{m}$. Thus,
\begin{equation*}
  P_\theta\left(\theta_n\init \ge t_{j\opt} + \frac{2 n \epsilon}{m}
  \right)
  \le \left(1 - \frac{n \epsilon}{m}\right)^\frac{n}{m}
  \le \exp\left(-\frac{n^2\epsilon}{m^2}\right).
\end{equation*}
Setting the number of bins $m = \sqrt{n}$,
the resolution $\epsilon = 2 \log n / n$, we obtain
$P_\theta(\theta_n\init \ge t_{j\opt} + 4 \log n / \sqrt{n})
\le n^{-2}$. Thus we have
\begin{equation}
  \label{eqn:quality-of-initial-estimate}
  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^2}.
\end{equation}

The second stage estimator follows roughly the same strategy, except that
the resolution of the bins is tighter. In particular, let us assume that
$|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$, which happens
eventually by inequality~\eqref{eqn:quality-of-initial-estimate}.  (We will
assume this tacitly for the remainder of the argument.)  Consider the
interval $\Theta_n \defeq \theta_n\init + [-\frac{16 \log n}{\sqrt{n}},
  \frac{16 \log n}{\sqrt{n}}]$ centered at $\theta_n\init$; we know that the
interval includes $[\theta - \frac{8 \log n}{\sqrt{n}}, \theta + \frac{8
    \log n}{\sqrt{n}}]$. Without loss of generality we assume $\theta_n\init
= 0$.  Following precisely the same discretization strategy as that for
$\theta_n\init$, we divide $\Theta_n$ into $m$ equal intervals, with
thresholds $t_j = -\frac{16 \log n}{\sqrt{n}} + \frac{32 (j - 1) \log n}{m
  \sqrt{n}}$; let $\epsilon_n = \frac{32 \log n}{m \sqrt{n}}$ be the width of
these intervals.  Then following exactly the same reasoning as above, we
assign indices $I_j = \{\frac{n(j - 1)}{m} + 1, \ldots, \frac{n j}{m}\}$ and
for $i \in I_j$, set $B_i = 1$ if $X_i \ge t_j$. We define $\theta_n$ to be
the maximal threshold $t_j$ for which $B_i = 0$ for all observations $X_i
\in I_j$. Then following precisely the reasoning above, we have (on the
event that $|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$)
\begin{equation*}
  P_\theta(|\theta_n - \theta|
  \ge 2 \epsilon_n)
  \le (1 - \epsilon_n)^\frac{n}{m}
  \le \exp\left(-\frac{n \epsilon_n}{m}\right)
  = \exp\left(-\frac{32 \sqrt{n} \log n}{m^2}\right).
\end{equation*}
Set $m = 4 n^{1/4}$ to obtain the claimed
result~\eqref{eqn:uniforms-are-easy}.

### 2019-09-30 preliminaries.tex
Trying to understand proofs

\section{Consistent Estimation and Off-the-shelf Bounds \label{sec:preliminary}}
%
In this section we adapt known results to derive lower and upper bounds on the relative efficiency under setting (iii). These bounds establish the following facts:
\begin{itemize}
\item[I.] A consistent estimator with an asymptotically normal distribution always exists in setting (iii), and hence in setting (ii). 
\item[II.] For the normal distribution, the ARE in setting (iii) is at most $3/4$. Namely, under setting (iii), all estimators are strictly inferior compared to the sample mean. 
\end{itemize}

\subsection{Consistent Estimation}
Fix $\theta_0 \in \mathbb R$ and define the $i$th message by 
\[
B_i = \mathbf 1_{X_i>\theta_0}, 
\]
where $\mathbf 1_A$ is the indicator of the event $A$. We have
\[
P_n \triangleq\frac{1}{n} \sum_{i=1}^n B_i \overset{a.s.}{\rightarrow} F(\theta - \theta_0),  
\]
so that 
\begin{equation}
\label{eq:estimator_naive}
{\theta}_n = \theta_0 + F^{-1}\left( P_n \right)
\end{equation}
is a consistent estimator for $\theta$ in the distributed setting of Figure~\ref{fig:setup}-(iii), where we note that $F(x)$ is invertible over the support of $f(x)$ which is a connected set due to log-concavity. Furthermore, the variance of $P_n$ is $F(\theta-\theta_0)\left(1-F(\theta-\theta_0)\right)$, and hence the delta method implies that ${\theta}_n$ is asymptotically normal with variance 
\begin{equation} 
 \frac{F(\theta-\theta_0)\left(1-F(\theta-\theta_0)\right)}{f^2(\theta-\theta_0)} = \frac{1}{
\eta(\theta-\theta_0)}.
\end{equation}
In particular, the ARE of ${\theta}_n$ equals $\eta(\theta - \theta_0)\sigma^2$. In other words, for a prescribed accuracy, ${\theta}_n$ of  \eqref{eq:estimator_naive} estimates $\theta$ with sample size that is $\eta(\theta- \theta_0)  \sigma^2$ times the samples size required for the sample mean. \par
%
%
Assumption~\ref{assump:failure_rate} implies that for  all $n$ large enough the ARE of ${\theta}_n$ of \eqref{eq:estimator_naive} is never greater than $\eta(0)$, which is the ARE of the sample median. This ARE is attained only when $\theta_0 = \theta$, whereas $\theta$ is apriori unknown. Since $\eta(x)$ vanishes as $|x|\rightarrow \infty$, the ARE of ${\theta}_n$ may be very small when $\theta$ is away from $\theta_0$. As an example, when $f(x)$ is a normal density, the ARE of ${\theta}_n$ is less than $\approx 0.14$ when $\theta_0$ is $2$ standard deviations from $\theta$. Therefore, the estimator ${\theta}_n$ has little practical value unless the radius of $\Theta$ is small compared to the standard deviation. \par 
It is suggested that we can attain lower variance in estimation by updating the threshold value $\theta_0$ in \eqref{eq:estimator_naive} after observing one or a batch of the single bit messages. Schemes with such adaptation fall within the adaptive setting of Figure~\ref{fig:setup}-(ii) which we consider in Section~\ref{sec:sequential}. \par

\subsection{Multiterminal Source Coding \label{sec:ceo}}
The CEO setting considers the estimation of a sequence $\theta_1,\theta_2\ldots$, where a noisy version of each $\theta_i$ is available at $n$ terminals. At each terminal $i$, an encoder observes the $k$ noisy samples
\[
X_{i,j} = \theta_j + Z_{i,j},\qquad j=1,\ldots,k, \qquad i = 1,\ldots,n,
\]
and transmits $r_i k$ bits to a central estimator  \cite{berger1996ceo}. The central estimator produces an estimates ${\theta}_1,\ldots,{\theta}_k$ with the goal of minimizing the quadratic risk:
\[
R_{\CEO} = \frac{1}{k} \sum_{j=1}^k \mathbb E \left[\left(\theta_j - {\theta}_j \right)^2 \right]. 
\]
Note that any distributed encoding scheme using one-bit per sample can be replicated $k$ times and thus leads to a legitimate encoding and estimation scheme for the CEO problem with $r_1=\ldots=r_n = 1$. It follows that, assuming that $\theta$ is drawn once from the prior $\pi(d\theta)$, our mean estimation problem from one-bit samples under distributed encoding corresponds to the CEO setting with $k=1$ realization of $\theta$ observed under noise at $n$ different locations, and communicated at each location using an encoder sending a single bit. 
%
Consequently, a lower bound on the MSE in estimating $\theta$ in the distributed encoding setting is given by the minimal MSE in the CEO setting as $k \to \infty$. Note that the difference between the CEO setting and ours lays in the privilege of each of the encoders to describe $k$ realizations of $\theta$ using $k$ bits with MSE averaged over these realizations, rather than a single realization using a single bit in ours. 
 \par
When the prior on $\theta$ and the noise corrupting it at each location are Gaussian, the optimal encoding scheme and its asymptotic risk as $k$ goes to infinity were fully characterized in \cite{prabhakaran2004rate}. Furthermore, the work of \cite{chen2004upper} provided an expression for quadratic risk attained in the CEO setting under Gaussian priors. Adapting to our setting, this expression provides the following proposition:
\begin{prop} \label{prop:ceo_lower_bound}
Assume that $\Theta = \mathbb R$ and $\pi(\theta) = \Ncal(0,\sigma_\theta^2)$ where $\sigma_\theta^2 \in \mathbb R$ is arbitrary. Then any estimator ${\theta}_n$ of $\theta$ in the distributed setting satisfies
\begin{equation} \label{eq:ceo_bound}
 n \ex{\left( \theta - \theta_n \right)^2} \geq \frac{4}{3} \sigma^2 + O(n^{-1}),
\end{equation}
where the expectation is with respect to $\theta$ and $X^n$.
\end{prop}

\begin{proof}
See Appendix~\ref{app:proof:prop_ceo_lower_bound}.
\end{proof}

From the formulation of the CEO problem, it follows that the difference between the MSE lower bound \eqref{eq:ceo_bound} and the actual MSE in the distributed setting (case (iii)) is attributed to the ability to perform coding over blocks. Namely, each encoder in the CEO setting may encode an arbitrary number of $k$ independent realizations of $\theta$ using $k$ bits, versus only one realization with one bit in ours. In other words, it is the ability to exploit the geometry of a high-dimensional product probability space that distinguishes between the CEO problem with one bit per encoder on average and the mean estimation problem from one-bit measurements in the distributed setting.

>>>>

\section{Consistent Estimation and Off-the-shelf Bounds \label{sec:preliminary}}

We begin our technical treatment by deriving a few
bounds on the efficiency of estimators in
setting~\eqref{item:distributed}. These bounds establish the
following facts:
\begin{enumerate}[1.]
\item A consistent estimator with an asymptotically normal distribution
  always exists in setting~\eqref{item:distributed}, and hence in the
  adaptive settings~\eqref{item:adaptive} and
  (\ref{item:one-step-adaptive}').
\item For the normal distribution, the asymptotic relative
  efficiency~\eqref{eqn:are-def} in the distributed
  setting~\eqref{item:distributed} is at most $3/4$. No estimator can be as
  efficient as the sample mean.
\end{enumerate}

\subsection{Consistent Estimation}
The simplest estimator is simply to invert a quantile. Indeed,
fix $\theta_0 \in \mathbb R$ and define the $i$th message by 
\[
B_i = \mathbf 1_{X_i>\theta_0}, 
\]
where $\mathbf 1_A$ is the indicator of the event $A$. We have
\[
P_n \triangleq\frac{1}{n} \sum_{i=1}^n B_i \overset{a.s.}{\rightarrow} F(\theta - \theta_0),  
\]
so that 
\begin{equation}
\label{eq:estimator_naive}
{\theta}_n = \theta_0 + F^{-1}\left( P_n \right)
\end{equation}
is a consistent estimator for $\theta$ in the distributed setting of
Figure~\ref{fig:setup}-(iii), where we note that $F$ is invertible over the
support of $f$. As the variance of $P_n$ is
$F(\theta-\theta_0)\left(1-F(\theta-\theta_0)\right)$, a delta method
calculation~\cite[Ch.~23]{VanDerVaart98} implies that ${\theta}_n$ is
asymptotically normal with variance
\begin{equation*}
  \frac{F(\theta-\theta_0)\left(1-F(\theta-\theta_0)\right)}{f^2(\theta-\theta_0)} = \frac{1}{
\eta(\theta-\theta_0)}.
\end{equation*}
In the Gaussian case where the $X_i \simiid \normal(\theta, \sigma^2)$, the
ARE of ${\theta}_n$ is $\eta(\theta - \theta_0)\sigma^2$.

Assumption~\ref{assump:failure_rate} implies that the optimal asymptotic
variance for an estimator of the form~\eqref{eq:estimator_naive} is $1 /
\eta(0)$, the asymptotic of the sample median. Unfortunately, as $\theta$ is
(by definition) \emph{a priori} unknown and $\eta(x)$ monotonically
decreases in $|x|$, this naive estimator $\theta_n$ may be very inefficient
when $\theta$ is far from the initial guess $\theta_0$. As an example, when
$f$ is a the normal density, the ARE of ${\theta}_n$ is less than $0.15$
when $|\theta_0 - \theta| \ge 2\sigma$, and more broadly, $\ARE(\theta_n)$
asymptotes to $|\theta_0| \exp(-\theta_0^2 / 2) / \sqrt{2\pi}$ as $|\theta_0
- \theta|$ gets large.  Yet that $\theta_0 = \theta$ minimizes this
asymptotic variance, and $\eta$ is continuous, is suggestive: if we can use
a suitably good initial estimate $\theta_n\init$ for $\theta$, it is
possible that a one-step adaptive estimator
(recall~(\ref{item:one-step-adaptive}')) may be asymptotically strong, as we
see in Section~\ref{sec:sequential}.

\subsection{Multiterminal Source Coding}
\label{sec:ceo}

A related problem is the CEO problem, which considers the estimation of a
sequence $\theta_1,\theta_2\ldots$, where a noisy version of each $\theta_i$
is available at $n$ terminals. At each terminal $i$, an encoder observes the
$k$ noisy samples
\[
X_{i,j} = \theta_j + Z_{i,j},\qquad j=1,\ldots,k, \qquad i = 1,\ldots,n,
\]
and transmits $r_i k$ bits to a central estimator~\cite{berger1996ceo}. The
central estimator produces an estimates ${\theta}_1,\ldots,{\theta}_k$ with
the goal of minimizing the quadratic risk:
\[
R_{\CEO} = \frac{1}{k} \sum_{j=1}^k \mathbb E \left[\left(\theta_j - {\theta}_j \right)^2 \right]. 
\]
Note that any distributed encoding scheme using one-bit per sample can be replicated $k$ times and thus leads to a legitimate encoding and estimation scheme for the CEO problem with $r_1=\ldots=r_n = 1$. It follows that, assuming that $\theta$ is drawn once from the prior $\pi(d\theta)$, our mean estimation problem from one-bit samples under distributed encoding corresponds to the CEO setting with $k=1$ realization of $\theta$ observed under noise at $n$ different locations, and communicated at each location using an encoder sending a single bit. 
%
Consequently, a lower bound on the MSE in estimating $\theta$ in the
distributed encoding setting is given by the minimal MSE in the CEO setting
as $k \to \infty$. Note that the difference between the CEO setting and ours
lays in the privilege of each of the encoders to describe $k$ realizations
of $\theta$ using $k$ bits with MSE averaged over these realizations, rather
than a single realization using a single bit in ours.

When the prior on $\theta$ and the noise corrupting it at each location are
Gaussian, Prabhakaran et al.~\cite{prabhakaran2004rate} characterize the
optimal encoding and its asymptotic risk as $k \to \infty$.  Chen et
al.~\cite{chen2004upper} also provide an expression for the quadratic risk
in the CEO setting under Gaussian priors. Adapting to our setting, this
expression provides the following proposition:
\begin{prop} \label{prop:ceo_lower_bound}
  Assume that $\Theta = \mathbb R$ and $\pi(\theta) =
  \Ncal(0,\sigma_\theta^2)$ where $\sigma_\theta^2 \in \mathbb R$ is
  arbitrary. Then any estimator ${\theta}_n$ of $\theta$ in the distributed
  setting satisfies
  \begin{equation} \label{eq:ceo_bound}
    n \ex{\left( \theta - \theta_n \right)^2} \geq \frac{4}{3} \sigma^2 + O(n^{-1}),
  \end{equation}
  where the expectation is with respect to $\theta$ and $X^n$.
\end{prop}
\noindent
See Appendix~\ref{app:proof:prop_ceo_lower_bound} for a proof.

As we shall see, this bound is loose: the difference between the MSE lower
bound~\eqref{eq:ceo_bound} and the actual MSE in the distributed setting
(case~\eqref{item:distributed}) occurs because in the CEO setting, each
encoder may encode an arbitrary number of $k$ independent realizations of
$\theta$ using $k$ bits; in our situation, $k = 1$. That blocking allows
more efficient encoding and exploiting the high-dimensional geometry of the
product probability space in the CEO problem is perhaps unsurprising, and
our goal in the sequel will be to characterize the performance degradation
one bit encoding engenders.
